---
title: "Validating bertscore() against the reference implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating bertscore() against the reference implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`ditto`'s `bertscore()` is a from-scratch reimplementation of the metric
introduced by Zhang et al. (2020). The authors also released a reference Python
package, [`bert-score`](https://github.com/Tiiiger/bert_score). This vignette
checks that `ditto` agrees with that reference: given the *same* model and the
*same* settings, the two should return the same precision, recall, and F1 for
every pair, up to small numerical differences.

## What has to match

The two implementations agree only when they compute the same thing, which holds
under five conditions. The run below satisfies all five:

1. **The same model** on both sides -- `bge-m3` here, served to `ditto` through
   `llama.cpp` and passed to `bert-score` via `model_type`. Any model both tools
   can load works (e.g. `roberta-large`, the `bert-score` default).

2. **The same layer.** A transformer emits a separate embedding for each token at
   every one of its layers, and BERTScore must pick one layer to score from. A
   `llama.cpp` `/embeddings` response returns the final layer, but `bert-score`
   defaults to a tuned intermediate one (layer 17 of 24 for `roberta-large`). That
   default is tuned for masked-language-model encoders, whose top layers drift
   toward the pretraining objective and lose general meaning; an intermediate
   layer correlates better with human judgment. `bge-m3` is a dedicated embedding
   model, though: it is trained so that its *final* layer is the semantic
   representation, so reading the last layer is the right choice rather than a
   compromise (and `bert-score` has no tuned layer for it anyway). The run sets
   `num_layers` to the model's last layer (24 for the large XLM-R behind `bge-m3`)
   so both read the final layer.

3. **No IDF weighting** -- `bertscore()` is unweighted, matched by `bert-score`'s
   `idf=False` (its default).

4. **No baseline rescaling** -- `bertscore()` returns raw scores absent a
   `baseline`, matched by `rescale_with_baseline=False` (its default) and no
   `baseline` on the `ditto` side.

5. **Identical input strings.** Both tools embed the raw string and neither
   lowercases, so the same text goes to each, uncleaned. The shared model means a
   shared tokenizer, so the token sequences and dropped special tokens line up.

## Test pairs

A spread of pairs exercises the metric across its range: a near-identical
rewording, a meaning-preserving paraphrase, a topically related pair, and an
unrelated pair. The point is not the absolute scores but that the two
implementations agree on them.

```{r}
#| label: pairs
candidates <- c(
  "to what extent do you agree with the statement",
  "how much do you agree with the statement",
  "how satisfied are you with the service",
  "what is your highest level of education"
)
references <- rep("to what extent do you agree with the statement", 4)
```

## ditto scores

`bertscore()` takes one pair at a time, mapped over the vectors:

```{r}
#| label: ditto-scores
#| eval: false
library(ditto)

# The GGUF build of BAAI/bge-m3 -- the same model bert-score loads via model_type.
start_llama_server(model = "bge-m3-f16.gguf")

ditto_scores <- t(mapply(bertscore, candidates, references, USE.NAMES = FALSE))

stop_llama_server()
```

| precision |  recall |      f1 |
|----------:|--------:|--------:|
|   1.00000 | 1.00000 | 1.00000 |
|   0.96706 | 0.96329 | 0.96517 |
|   0.79389 | 0.79100 | 0.79244 |
|   0.72340 | 0.71281 | 0.71807 |

## Reference scores

The same pairs through `bert-score` with the settings from *What has to match*:

```python
from bert_score import score

candidates = [
    "to what extent do you agree with the statement",
    "how much do you agree with the statement",
    "how satisfied are you with the service",
    "what is your highest level of education",
]
references = ["to what extent do you agree with the statement"] * 4

P, R, F = score(
    candidates,
    references,
    model_type="BAAI/bge-m3",
    num_layers=24,                 # final layer, to match llama.cpp
    idf=False,                     # unweighted, to match bertscore()
    rescale_with_baseline=False,   # raw scores, to match bertscore()
)
```

| precision |  recall |      f1 |
|----------:|--------:|--------:|
|   1.00000 | 1.00000 | 1.00000 |
|   0.96709 | 0.96333 | 0.96521 |
|   0.79418 | 0.79127 | 0.79272 |
|   0.72388 | 0.71323 | 0.71852 |

## Comparing the two

The two implementations agree score for score; nowhere do they differ by more
than 0.0005. That confirms `ditto` computes the same metric as the reference.

| precision |  recall |      f1 |
|----------:|--------:|--------:|
|   0.00000 | 0.00000 | 0.00000 |
|   0.00003 | 0.00004 | 0.00004 |
|   0.00029 | 0.00027 | 0.00028 |
|   0.00048 | 0.00042 | 0.00045 |

## References

Zhang, T., Kishore, V., Wu, F., Weinberger, K. Q., & Artzi, Y. (2020).
BERTScore: Evaluating text generation with BERT. *International Conference on
Learning Representations (ICLR)*.
<https://doi.org/10.48550/arXiv.1904.09675>
