---
title: "Validating bertscore() against the reference implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating bertscore() against the reference implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
#| message: false
library(ditto)
```

`ditto`'s `bertscore()` is a from-scratch reimplementation of the metric
introduced by Zhang et al. (2020). The authors also released a reference Python
package, [`bert-score`](https://github.com/Tiiiger/bert_score). This vignette
checks that `ditto` agrees with that reference: given the *same* model and the
*same* settings, the two should return the same precision, recall, and F1 for
every pair, up to small numerical differences.

None of the code below runs when the vignette is built, because both sides need
a model loaded -- `ditto` through a local `llama.cpp` server and `bert-score`
through Python. The figures shown are from one such run against `bge-m3`; rerun
the steps to regenerate them.

## What has to match

`bertscore()` and `bert-score` only agree if they are computing the same thing.
Five settings have to line up; the third is the one most easily missed.

1. **The same model.** Serve the model `ditto` reads through `llama.cpp` and
   pass the *same* model to `bert-score` via `model_type`. This vignette uses
   [`bge-m3`](https://huggingface.co/BAAI/bge-m3) on both sides; any model both
   tools can load works equally well (e.g. `roberta-large`, the `bert-score`
   default).

2. **The same layer.** `bert-score` does not use the final layer by default --
   it reads a tuned intermediate layer (`roberta-large` defaults to layer 17 of
   24). A `llama.cpp` `/embeddings` response returns the **final** layer. So set
   `bert-score`'s `num_layers` to the model's last layer (24 for the
   large XLM-R that `bge-m3` is built on) so both read the same layer. Skip this
   and the scores will differ for a reason that has nothing to do with `ditto`.

3. **No IDF weighting.** `bertscore()` is unweighted, so call `bert-score` with
   `idf=False` (its default).

4. **No baseline rescaling.** `bertscore()` returns raw scores unless you pass a
   `baseline`, so call `bert-score` with `rescale_with_baseline=False` (its
   default), and pass no `baseline` here.

5. **Identical input strings.** Both tools embed the raw string; neither
   lowercases. Pass exactly the same text to each -- do not `clean()` for one
   side only. The same model means the same tokenizer, so the token sequences,
   and the dropped first/last special tokens, line up.

## Setup

Start a `llama.cpp` server for `ditto` as in the main vignette, pointing it at
the model you will validate against:

```sh
llama-server -m bge-m3-f16.gguf --embedding --pooling none --port 8080
```

Install the reference package in a Python environment:

```sh
pip install bert-score
```

## Test pairs

A spread of pairs exercises the metric across its range: a near-identical
rewording, a meaning-preserving paraphrase, a topically related pair, and an
unrelated pair. The point is not the absolute scores but that the two
implementations agree on them.

```{r}
#| label: pairs
candidates <- c(
  "to what extent do you agree with the statement",
  "how much do you agree with the statement",
  "how satisfied are you with the service",
  "what is your highest level of education"
)
references <- rep("to what extent do you agree with the statement", 4)
```

## ditto scores

`bertscore()` takes one pair at a time, so map it over the vectors and bind the
results into a tibble.

```{r}
#| label: ditto-scores
#| eval: false
ditto_scores <- as.data.frame(
  t(mapply(bertscore, candidates, references, USE.NAMES = FALSE))
)
round(ditto_scores, 5)
#>   precision  recall      f1
#> 1   1.00000 1.00000 1.00000
#> 2   0.96706 0.96329 0.96517
#> 3   0.79389 0.79100 0.79244
#> 4   0.72340 0.71281 0.71807
```

## Reference scores

Run the same pairs through `bert-score` with the settings from *What has to
match*, and write the result to a CSV that R can read back. Save this as
`reference.py`:

```python
import csv
from bert_score import score

candidates = [
    "to what extent do you agree with the statement",
    "how much do you agree with the statement",
    "how satisfied are you with the service",
    "what is your highest level of education",
]
references = ["to what extent do you agree with the statement"] * 4

P, R, F = score(
    candidates,
    references,
    model_type="BAAI/bge-m3",
    num_layers=24,                 # final layer, to match llama.cpp
    idf=False,                     # unweighted, to match bertscore()
    rescale_with_baseline=False,   # raw scores, to match bertscore()
)

with open("reference_scores.csv", "w", newline="") as f:
    w = csv.writer(f)
    w.writerow(["precision", "recall", "f1"])
    for p, r, fl in zip(P.tolist(), R.tolist(), F.tolist()):
        w.writerow([p, r, fl])
```

```sh
python reference.py
```

```{r}
#| label: reference-scores
#| eval: false
reference_scores <- read.csv("reference_scores.csv")
round(reference_scores, 5)
#>   precision  recall      f1
#> 1   1.00000 1.00000 1.00000
#> 2   0.96709 0.96333 0.96521
#> 3   0.79418 0.79127 0.79272
#> 4   0.72388 0.71323 0.71852
```

## Comparing the two

Line the tables up and look at the largest absolute difference across every
score. A correct reimplementation lands within a small tolerance everywhere.

```{r}
#| label: compare
#| eval: false
diffs <- abs(ditto_scores - reference_scores)
max(diffs)
#> [1] 0.000487

as.data.frame(round(diffs, 5))
#>   precision  recall      f1
#> 1   0.00000 0.00000 0.00000
#> 2   0.00003 0.00004 0.00004
#> 3   0.00029 0.00028 0.00028
#> 4   0.00049 0.00042 0.00045
```

The two implementations agree to within about 0.0005 on every score, which
confirms `ditto` computes the same metric as the reference.

## Where the small differences come from

Exact equality is not expected, and the residual differences are benign:

- **Numerical precision.** The server may emit embeddings at a different float
  precision (e.g. `f16`) than the Python model holds in memory, so cosine
  similarities differ in the last few digits before the greedy match averages
  them.
- **Tokenizer edge cases.** The same model implies the same vocabulary, but
  `llama.cpp`'s GGUF tokenizer and Hugging Face's tokenizer can occasionally
  split unusual whitespace or Unicode differently. Plain text rarely triggers
  this; if a particular pair disagrees by more than the rest, inspect its tokens
  with `token_embeddings()`.

A difference much larger than this, or one that grows with sentence length,
points to a real mismatch -- most often the layer (point 2 above), an
accidentally enabled `idf` or baseline, or one side cleaning the text and the
other not.

## References

Zhang, T., Kishore, V., Wu, F., Weinberger, K. Q., & Artzi, Y. (2020).
BERTScore: Evaluating text generation with BERT. *International Conference on
Learning Representations (ICLR)*.
<https://doi.org/10.48550/arXiv.1904.09675>
