---
title: "Comparing strings with ditto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing strings with ditto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
#| message: false
library(ditto)
```

`ditto` compares two pieces of text and reduces the comparison to a single
number, so that comparisons can be ranked, filtered, or summarised at scale.
Typical uses are checking whether a survey question has been reworded across
waves, comparing a translation against its source, and grouping near-duplicate
free-text responses.

The metrics differ in what counts as a difference. Three character-level
metrics -- edit distance, Jaccard, and cosine -- come from
[`stringdist`](https://github.com/markvanderloo/stringdist) and are available
only through `compare_strings()`. Six word-level metrics -- `bleu()`, `chrf()`,
`rouge()`, `ter()`, `wer()`, and `meteor()` -- compare word or character
sequences, and the embedding metrics `bertscore()` and `cosine_similarity()`
compare meaning through contextual embeddings; all of these can also be called
on their own. `compare_strings()` runs the character- and word-level metrics
together, and the embedding metrics too when an embedding server is available,
returning them in one table.

## Cleaning text

Two strings can differ only in capitalisation or punctuation while expressing
exactly the same thing. `clean()` removes those surface differences by
lowercasing, stripping punctuation, and collapsing repeated whitespace, so
that they do not lower the similarity scores.

```{r}
#| label: clean
clean("To what extent do you AGREE with the statement?")
```

The rest of this vignette compares a reference survey question against four
variants that range from a near-identical rewording to an unrelated question.
Each string is cleaned first.

```{r}
#| label: data
reference <- clean("To what extent do you agree with the following statements?")

variants <- c(
  "To what extent do you agree with the following statement?",
  "How much do you agree with each of the statements below?",
  "How satisfied are you with the service you received?",
  "What is your highest level of education completed?"
)
labels <- c("Near-identical", "Paraphrase", "Different", "Unrelated")
variants <- clean(variants)
```

## Character metrics

`compare_strings()` computes every non-embedding metric for each candidate
against the reference and returns a tibble with one row per pair. Three of the
columns are character metrics from `stringdist`: Levenshtein (single-character
edits), Jaccard (overlap of the character sets), and cosine (overlap of
character frequencies).

```{r}
#| label: compare
result <- compare_strings(variants, reference)
character_metrics <- c("levenshtein", "jaccard", "cosine")
knitr::kable(cbind(label = labels, round(result[character_metrics], 2)))
```

The near-identical variant scores near 1 on all three. The other variants
diverge in ways that reveal what each metric responds to. The unrelated
question still scores 0.92 on cosine -- here a *character* cosine, the overlap
of character frequencies, not to be confused with the meaning-based
`cosine_similarity()` introduced later -- because it happens to reuse many of
the same characters. None of these metrics compares meaning, so surface overlap
can make an unrelated string look similar.

## BLEU

`bleu()` evaluates a candidate against a reference at the level of word
sequences rather than characters. It counts how many of the candidate's
n-grams (single words, pairs, triples, and 4-word sequences) also appear in
the reference, credits each reference n-gram only as many times as it occurs,
and applies a penalty when the candidate is shorter than the reference. The
maximum n-gram order is capped at the length of the shorter string, so short
inputs are scored on the orders that are actually possible.

```{r}
#| label: bleu
bleu(
  "how much do you agree with the statement",
  "to what extent do you agree with the statement"
)
```

BLEU returns 0 when any n-gram order has no matches at all, which is why the
"Different" and "Unrelated" variants score 0 in the table below: neither
shares a 4-word sequence with the reference. This makes BLEU stricter than the
character metrics on word order, and well suited to translation evaluation,
where the arrangement of words carries meaning.

## Beyond BLEU: CHRF, ROUGE, TER, and METEOR

BLEU's all-or-nothing behaviour, and its indifference to recall, motivated four
later metrics. Each is available on its own and as a column of
`compare_strings()`.

`chrf()` counts *character* n-grams rather than word n-grams, and combines
precision with recall. Because it works below the word, an inflected form that
shares no whole word with the reference still shares most of its characters:

```{r}
#| label: chrf
c(chrf = chrf("agreeing", "agreed"), bleu = bleu("agreeing", "agreed"))
```

Whitespace is stripped before the character n-grams are extracted, so spacing
cannot change the score. `chrf()` agrees with `sacrebleu`'s CHRF to within
floating-point error; `vignette("metrics")` records what every metric here is
validated against.

`rouge()` is the recall-oriented counterpart to BLEU. Variants `"1"` and `"2"`
count shared unigrams and bigrams; variant `"l"` matches on the longest common
subsequence of words, so a long run of words in the same order scores highly
even when other words are interspersed. There is no brevity penalty, so a short
candidate is not punished for what it omits beyond the loss of recall.

```{r}
#| label: rouge
rouge("the cat sat on the mat", "a cat was sitting on the mat")
rouge("the cat sat on the mat", "a cat was sitting on the mat", variant = "l")
```

`ter()` reports Translation Edit Rate: the number of edits needed to turn the
candidate into the reference, divided by the reference length. It is an *error*
rate rather than a similarity, so 0 is a perfect match, lower is better, and it
is not bounded above by 1. Alongside insertions, deletions, and substitutions it
counts a *shift* -- a contiguous block of words moved elsewhere -- as a single
edit, however many words the block holds.

`wer()` is the same rate without shifts, which is exactly word error rate. The
pair is more informative than either alone: where they diverge, the candidate
has the right words in the wrong place.

```{r}
#| label: ter
ter("the cat sat on the mat", "a cat was sitting on the mat")

# A block of three words moved: one shift, or six rewrites.
c(ter = ter("d e f a b c", "a b c d e f"), wer = wer("d e f a b c", "a b c d e f"))
```

`meteor()` matches words by stem as well as by exact form, weights recall nine
times as heavily as precision, and then discounts the result by a fragmentation
penalty counting how many contiguous chunks the matched words fall into.
Reordered words are still matched -- the penalty, not a lost match, is what
charges for the reordering:

```{r}
#| label: meteor
meteor("the cats are agreeing", "the cat agreed")
meteor("the cat sat", "sat the cat") # every word matches, across two chunks
```

Synonym matching, METEOR's third stage, is not implemented: it needs a WordNet
installation that exists for only a handful of languages. On text where no
synonyms would have matched, `meteor()` agrees with `nltk`'s implementation to
within floating-point error.

Stemming is language-specific, so pass `language` for text that is not English.
Left at the default, the English stemmer sees no match between Dutch plurals
and their singulars:

```{r}
#| label: meteor-language
c(
  dutch = meteor("de katten liepen", "de kat liep", language = "dutch"),
  english = meteor("de katten liepen", "de kat liep")
)
```

Across the four variants these metrics separate the near-identical rewording
from the unrelated question far more sharply than the character metrics did:

```{r}
#| label: word-metrics
word_metrics <- c("bleu", "chrf", "rouge_1", "rouge_l", "ter", "wer", "meteor")
knitr::kable(cbind(label = labels, round(result[word_metrics], 2)))
```

BLEU collapses to 0 for both the "Different" and "Unrelated" variants, which
tells you only that neither shares a 4-gram. The others keep ranking them:
METEOR falls from 0.26 to 0.05 and TER climbs from 0.80 to 0.90. Two columns
agree throughout, each for its own reason. `rouge_1` and `rouge_l` match because
the words these variants share with the reference happen to appear in the same
order. `ter` and `wer` match because no block of words is merely displaced here;
they part company on a candidate that reorders the reference rather than
rewording it.

## Semantic similarity with embeddings

The metrics so far compare the characters or words that are present. A
paraphrase that conveys the same meaning with different words scores low on all
of them. The package's two embedding metrics compare *meaning* instead.
`bertscore()` embeds each token in the context of its sentence and matches every
candidate token to its most similar reference token, reporting precision,
recall, and their harmonic mean (F1). `cosine_similarity()` instead pools each
string into a single vector and takes the cosine between the two.

These are the only parts of the package that need an external model, because
they require one embedding *per token* rather than one pooled vector per string.

### Setting up the embedding server

`bertscore()` and `cosine_similarity()` read per-token embeddings from a local
[`llama.cpp`](https://github.com/ggml-org/llama.cpp) server. The friendly
local runners, Ollama and LM Studio, return only a single pooled vector per
string through their OpenAI-style endpoints, so they cannot supply the
token-level embeddings these metrics need. A `llama.cpp` server started with
`--pooling none` can.

1. Download the prebuilt `llama.cpp` binaries for your platform from the
   [releases page](https://github.com/ggml-org/llama.cpp/releases) (no
   compilation required).
2. Download an **embedding model** in GGUF format -- one trained so its output is
   a semantic vector, which is what these metrics compare (a plain generative or
   masked-language model is not built for this). This vignette uses
   [`bge-m3`](https://huggingface.co/BAAI/bge-m3), a multilingual model that
   needs no task prefix; `all-MiniLM-L6-v2` is a smaller, English-only
   alternative. Some models, such as the E5 family, require a task prefix like
   `"query: "`; pass it through the `prefix` argument so its tokens are stripped
   before matching, rather than treated as part of the text.
3. Start the server with embeddings and pooling disabled:

```sh
llama-server -m bge-m3-f16.gguf --embedding --pooling none --port 8080
```

The default `host` argument points at `http://localhost:8080`; pass a
different URL if the server runs elsewhere.

Rather than start the server by hand, `start_llama_server()` launches it for
you with the right flags and waits until it is ready to answer requests.
`stop_llama_server()` shuts it down again, and `llama_server_running()` reports
whether a server is responding. Give it the paths to your `llama-server`
binary and `.gguf` model; `exe` can be omitted if `llama-server` is on your
`PATH`.

```{r}
#| label: start-server
#| eval: false
start_llama_server(
  model = "path/to/bge-m3-f16.gguf",
  exe = "path/to/llama-server"
)
llama_server_running()
```

If you set the binary and model paths once as options in your `.Rprofile`,
`start_llama_server()` needs no arguments:

```{r}
#| label: server-options
#| eval: false
options(
  ditto.llama_server = "path/to/llama-server",
  ditto.llama_model = "path/to/bge-m3-f16.gguf"
)
start_llama_server()
```

### Computing a BERTScore

With the server running, `bertscore()` returns a named vector of precision,
recall, and F1. The examples that follow are not run when this vignette is
built, since they depend on the server; the output shown is from a live run
against `bge-m3`.

```{r}
#| label: bertscore
#| eval: false
bertscore("how much do you agree", "to what extent do you agree")
bertscore("what is your date of birth", "to what extent do you agree")
```

| pair       | precision | recall |    f1 |
|:-----------|----------:|-------:|------:|
| paraphrase |     0.944 |  0.938 | 0.941 |
| unrelated  |     0.709 |  0.720 | 0.714 |

The paraphrase scores 0.94 and the unrelated question 0.71. Both are high:
`bge-m3` places even unrelated text at a substantial similarity, so the raw
separation is narrow. The ordering is correct, but these scores need calibrating
before they are comfortable as thresholds; *Calibrating scores with a baseline*
below does exactly that. `token_embeddings()` exposes the underlying matrix if
you want to inspect it directly; it has one row per token, after dropping the
model's special tokens, and one column per embedding dimension.

```{r}
#| label: token-embeddings
#| eval: false
dim(token_embeddings("to what extent do you agree"))
```

| rows (tokens) | columns (dimensions) |
|--------------:|---------------------:|
|             6 |                 1024 |

### Whole-string cosine

`cosine_similarity()` is the pooled counterpart to `bertscore()`: instead of
matching tokens, it collapses each string into a single vector and takes the
cosine between them. It is faster and is the classic sentence-embedding
similarity. Because the server returns per-token vectors, the pooling into one
vector happens in R, controlled by `pooling`. `"mean"` averages the token
vectors, as sentence-transformers models like `all-MiniLM` and the E5 family do;
`"cls"` takes the leading special token's vector, as BGE models like `bge-m3` do.
Use the pooling your model was trained with.

```{r}
#| label: cosine
#| eval: false
# bge-m3 is a CLS-pooling model.
cosine_similarity("how much do you agree", "to what extent do you agree",
                  pooling = "cls")
cosine_similarity("what is your date of birth", "to what extent do you agree",
                  pooling = "cls")
```

| pair       | cosine similarity |
|:-----------|------------------:|
| paraphrase |             0.892 |
| unrelated  |             0.474 |

With the matching CLS pooling the unrelated pair drops to 0.47 while the
paraphrase stays at 0.89, a wider gap than mean pooling gives (0.72 versus 0.95),
because the mean blurs a sentence into an average that even unrelated questions
share. This `cosine_similarity()` compares meaning and should not be confused
with the character-frequency `cosine` column of `compare_strings()`.

### Matching across languages

Because `bge-m3` is multilingual, the embedding metrics recognise a paraphrase
even across languages, where the character and word metrics see no overlap at
all. Comparing the English reference with its Dutch translation:

```{r}
#| label: multilingual
#| eval: false
bertscore("to what extent do you agree", "in hoeverre bent u het eens")[["f1"]]
cosine_similarity("to what extent do you agree", "in hoeverre bent u het eens",
                  pooling = "cls")
```

| metric            | score |
|:------------------|------:|
| BERTScore F1      | 0.921 |
| cosine similarity | 0.895 |

Both score the cross-lingual pair almost as high as an English paraphrase.

### All metrics together

Setting `bert = TRUE` adds the two embedding columns, `bertscore_f1` and
`cosine_emb`, to `compare_strings()`; pass `pooling = "cls"` for `bge-m3`.
Placing them beside the surface metrics makes the difference between surface
and semantic similarity visible. The call returns every column shown earlier;
a representative few are reproduced here.

```{r}
#| label: compare-bert
#| eval: false
compare_strings(variants, reference, bert = TRUE, pooling = "cls")
```

| label          | cosine | bleu | meteor |  ter | bertscore_f1 | cosine_emb |
|:---------------|-------:|-----:|-------:|-----:|-------------:|-----------:|
| Near-identical |   1.00 | 0.88 |   1.00 | 0.10 |         0.99 |       0.97 |
| Paraphrase     |   0.94 | 0.26 |   0.56 | 0.70 |         0.96 |       0.93 |
| Different      |   0.84 | 0.00 |   0.26 | 0.80 |         0.78 |       0.65 |
| Unrelated      |   0.92 | 0.00 |   0.05 | 0.90 |         0.71 |       0.49 |

Both embedding metrics keep the paraphrase high (0.96 and 0.93) even though its
BLEU is low (0.26) and its METEOR only 0.56, because it preserves meaning while
changing words. The different and unrelated questions, which character cosine
left at a misleading 0.84 and 0.92, separate much better on `cosine_emb` (0.65
and 0.49) than on the anisotropic `bertscore_f1` (0.78 and 0.71) -- though the
baseline rescaling below recovers that separation for BERTScore too.

### Calibrating scores with a baseline

Raw BERTScore values are not comparable across models, and some models compress
the range. With `all-MiniLM-L6-v2` an unrelated pair scores around 0.11, but
with a model whose token embeddings are anisotropic, such as the multilingual
`bge-m3`, *unrelated* text already scores around 0.71. The ordering is still
correct -- paraphrases score higher than unrelated pairs -- but the usable range
is squeezed into a narrow band near the top, and a fixed threshold means
something different for each model.

The original BERTScore addresses this by rescaling against a baseline: the score
the model assigns to unrelated text. `bertscore_baseline()` estimates it by
averaging over many random, and therefore unrelated, pairs drawn from a corpus.
By default it uses the English rows of `example_sentences`, a bundled set of
distinct-meaning sentences; pass your own `texts` for a domain- or
language-specific estimate.

```{r}
#| label: baseline
#| eval: false
baseline <- bertscore_baseline(seed = 1)
baseline
```

| precision | recall |    f1 |
|----------:|-------:|------:|
|     0.628 |  0.629 | 0.628 |

Passing this `baseline` to `bertscore()` (or `compare_strings()`) rescales each
score as `(x - baseline) / (1 - baseline)`, mapping the unrelated-text floor to
0 while leaving a perfect match at 1.

```{r}
#| label: baseline-rescale
#| eval: false
# Raw scores from bge-m3: paraphrase and unrelated sit close together.
bertscore("how much do you agree", "to what extent do you agree")[["f1"]]
bertscore("what is your date of birth", "to what extent do you agree")[["f1"]]

# Rescaled against the baseline: the gap opens up.
bertscore(
  "how much do you agree",
  "to what extent do you agree",
  baseline = baseline
)[["f1"]]
bertscore(
  "what is your date of birth",
  "to what extent do you agree",
  baseline = baseline
)[["f1"]]
```

| pair       | raw f1 | rescaled f1 |
|:-----------|-------:|------------:|
| paraphrase |  0.941 |       0.842 |
| unrelated  |  0.714 |       0.231 |

The baseline is specific to the model and the language, so estimate it from text
that matches your use. For a multilingual model, `bertscore_baselines()` returns
one baseline per language from a labelled corpus (`example_sentences` by
default), so each language can be rescaled against its own floor:

```{r}
#| label: baseline-language
#| eval: false
baselines <- bertscore_baselines(seed = 1)
bertscore("in hoeverre bent u het eens", "hoeveel bent u het eens",
          baseline = baselines[["nl"]])
```

## Choosing a metric

The metrics answer different questions, so the choice follows from what counts
as a difference.

- Use the **character metrics** when responses should closely match the
  reference and you expect only typos, small edits, or reordering. They are
  fast and need no external model. Cosine is usually the most informative of
  the three; Jaccard adds little beside it.
- Use **BLEU** when word order matters, such as translation evaluation, and
  you want a candidate penalised for rearranging or dropping words. Be aware
  that it collapses to 0 as soon as one n-gram order has no match, so it
  cannot rank the candidates that miss badly.
- Use **CHRF** for morphologically rich languages, or wherever inflection
  should not count as a mismatch. Because it scores character n-grams, it
  degrades gracefully instead of collapsing to 0.
- Use **ROUGE** when coverage of the reference matters more than precision, as
  in summarisation. Variant `"l"` is the forgiving one: it rewards a long run
  of words in the right order and ignores what is interspersed.
- Use **METEOR** when you want one number that credits stem matches, leans on
  recall, and still charges for reordering. Set `language` to match your text.
- Use **TER** when you want to report how much editing a candidate needs
  rather than how similar it is, and a displaced block of words should cost one
  edit rather than many. Remember it runs the other way: 0 is perfect, and it
  can exceed 1.
- Use **WER** where word order is not negotiable, as in speech recognition. It
  is `ter()` without the shift, so reading the two side by side tells you how
  much of the error is misplacement rather than wrong words.
- Use the **embedding metrics** when paraphrasing is expected and valid
  responses may use different words, or even different languages, for the same
  meaning. They are the only metrics here that score such responses correctly,
  at the cost of running a local embedding server. `cosine_similarity()` is
  faster and gives one number per pair; `bertscore()` is finer-grained, matching
  token by token, and reports precision and recall separately.

Two practical notes for the embedding metrics: match the `pooling` argument to
your model (`"cls"` for `bge-m3`, `"mean"` for `all-MiniLM` and the E5 family),
and rescale `bertscore()` against a baseline when you need thresholds that are
comparable across models.
