---
title: "Comparing strings with ditto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing strings with ditto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
#| message: false
library(ditto)
```

`ditto` compares two pieces of text and reduces the comparison to a single
number, so that comparisons can be ranked, filtered, or summarised at scale.
Typical uses are checking whether a survey question has been reworded across
waves, comparing a translation against its source, and grouping near-duplicate
free-text responses.

The package collects metrics that differ in what they treat as a meaningful
difference. The edit-distance, Jaccard, and cosine metrics come from
[`stringdist`](https://github.com/markvanderloo/stringdist) and compare
characters; `bleu()` compares word sequences; and `bertscore()` compares
meaning through contextual embeddings. `compare_strings()` runs the
character- and word-level metrics together and returns them in one table.

## Cleaning text

Two strings can differ only in capitalisation or punctuation while expressing
exactly the same thing. `clean()` removes those surface differences by
lowercasing, stripping punctuation, and collapsing repeated whitespace, so
that they do not lower the similarity scores.

```{r}
#| label: clean
clean("To what extent do you AGREE with the statement?")
```

The rest of this vignette compares a reference survey question against four
variants that range from a near-identical rewording to an unrelated question.
Each string is cleaned first.

```{r}
#| label: data
reference <- clean("To what extent do you agree with the following statements?")

variants <- c(
  "To what extent do you agree with the following statement?",
  "How much do you agree with each of the statements below?",
  "How satisfied are you with the service you received?",
  "What is your highest level of education completed?"
)
labels <- c("Near-identical", "Paraphrase", "Different", "Unrelated")
variants <- clean(variants)
```

## Character and word metrics

`compare_strings()` computes four metrics for each candidate against the
reference and returns a tibble with one row per pair. Three are character
metrics from `stringdist`: Levenshtein (single-character edits), Jaccard
(overlap of the character sets), and cosine (overlap of character
frequencies). The fourth is BLEU, which is described below.

```{r}
#| label: compare
result <- compare_strings(variants, reference)
cbind(label = labels, round(result[, -(1:2)], 2))
```

The near-identical variant scores near 1 on every metric. The other three
diverge in ways that reveal what each metric responds to. The unrelated
question still scores 0.92 on cosine, because it happens to reuse many of the
same characters. None of these metrics compares meaning, so surface overlap
can make an unrelated string look similar.

## BLEU

`bleu()` evaluates a candidate against a reference at the level of word
sequences rather than characters. It counts how many of the candidate's
n-grams (single words, pairs, triples, and 4-word sequences) also appear in
the reference, credits each reference n-gram only as many times as it occurs,
and applies a penalty when the candidate is shorter than the reference. The
maximum n-gram order is capped at the length of the shorter string, so short
inputs are scored on the orders that are actually possible.

```{r}
#| label: bleu
bleu(
  "how much do you agree with the statement",
  "to what extent do you agree with the statement"
)
```

BLEU returns 0 when any n-gram order has no matches at all, which is why the
"Different" and "Unrelated" variants score 0 in the table above: neither
shares a 4-word sequence with the reference. This makes BLEU stricter than the
character metrics on word order, and well suited to translation evaluation,
where the arrangement of words carries meaning.

## Semantic similarity with BERTScore

The metrics so far compare the characters or words that are present. A
paraphrase that conveys the same meaning with different words scores low on
all of them. `bertscore()` compares meaning instead. It embeds each token in
the context of its sentence, matches every candidate token to its most similar
reference token, and reports precision, recall, and their harmonic mean (F1).

This is the only part of the package that needs an external model, because it
requires one embedding *per token* rather than one pooled vector per string.

### Setting up the embedding server

`bertscore()` reads per-token embeddings from a local
[`llama.cpp`](https://github.com/ggml-org/llama.cpp) server. The friendly
local runners, Ollama and LM Studio, return only a single pooled vector per
string through their OpenAI-style endpoints, so they cannot supply the
token-level embeddings this metric needs. A `llama.cpp` server started with
`--pooling none` can.

1. Download the prebuilt `llama.cpp` binaries for your platform from the
   [releases page](https://github.com/ggml-org/llama.cpp/releases) (no
   compilation required).
2. Download an embedding model in GGUF format. `all-MiniLM-L6-v2` is small and
   works well here. Prefer a model without a required task prefix; models such
   as `nomic-embed-text` prepend tokens like `search_document:` that would be
   matched along with the real text.
3. Start the server with embeddings and pooling disabled:

```sh
llama-server -m all-MiniLM-L6-v2-f16.gguf --embedding --pooling none --port 8080
```

The default `host` argument points at `http://localhost:8080`; pass a
different URL if the server runs elsewhere.

Rather than start the server by hand, `start_llama_server()` launches it for
you with the right flags and waits until it is ready to answer requests.
`stop_llama_server()` shuts it down again, and `llama_server_running()` reports
whether a server is responding. Give it the paths to your `llama-server`
binary and `.gguf` model; `exe` can be omitted if `llama-server` is on your
`PATH`.

```{r}
#| label: start-server
#| eval: false
start_llama_server(
  model = "path/to/all-MiniLM-L6-v2-f16.gguf",
  exe = "path/to/llama-server"
)
llama_server_running()
```

If you set the binary and model paths once as options in your `.Rprofile`,
`start_llama_server()` needs no arguments:

```{r}
#| label: server-options
#| eval: false
options(
  ditto.llama_server = "path/to/llama-server",
  ditto.llama_model = "path/to/all-MiniLM-L6-v2-f16.gguf"
)
start_llama_server()
```

### Computing a score

With the server running, `bertscore()` returns a named vector of precision,
recall, and F1. The following examples are not run when this vignette is
built, since they depend on the server; the output shown is from a live run
against `all-MiniLM-L6-v2`.

```{r}
#| label: bertscore
#| eval: false
bertscore("how much do you agree", "to what extent do you agree")
#>  precision     recall         f1
#>      0.599      0.583      0.591

bertscore("what is your date of birth", "to what extent do you agree")
#>  precision     recall         f1
#>      0.114      0.102      0.108
```

The paraphrase scores 0.59 and the unrelated question 0.11. The separation
reflects meaning rather than shared characters. `token_embeddings()` exposes
the underlying matrix if you want to inspect it directly; it has one row per
token, after dropping the model's special tokens, and one column per embedding
dimension.

```{r}
#| label: token-embeddings
#| eval: false
dim(token_embeddings("to what extent do you agree"))
#> [1]   6 384
```

### BERTScore alongside the other metrics

Setting `bert = TRUE` adds a BERTScore F1 column to `compare_strings()`. The
full table makes the difference between surface and semantic similarity
visible. The numbers below are from a live run.

```{r}
#| label: compare-bert
#| eval: false
compare_strings(variants, reference, bert = TRUE)
```

| label          | levenshtein | jaccard | cosine | bleu | bertscore_f1 |
|:---------------|------------:|--------:|-------:|-----:|-------------:|
| Near-identical |        0.98 |    1.00 |   1.00 | 0.88 |         0.92 |
| Paraphrase     |        0.49 |    0.86 |   0.94 | 0.26 |         0.78 |
| Different      |        0.35 |    0.67 |   0.84 | 0.00 |         0.23 |
| Unrelated      |        0.25 |    0.82 |   0.92 | 0.00 |         0.17 |

The paraphrase keeps a high BERTScore (0.78) even though its BLEU is low
(0.26), because it preserves meaning while changing words. The unrelated
question drops to 0.17 on BERTScore, where cosine left it at a misleading
0.92.

## Choosing a metric

The metrics answer different questions, so the choice follows from what counts
as a difference.

- Use the **character metrics** when responses should closely match the
  reference and you expect only typos, small edits, or reordering. They are
  fast and need no external model. Cosine is usually the most informative of
  the three; Jaccard adds little beside it.
- Use **BLEU** when word order matters, such as translation evaluation, and
  you want a candidate penalised for rearranging or dropping words.
- Use **BERTScore** when paraphrasing is expected and valid responses may use
  different words for the same meaning. It is the only metric here that scores
  such responses correctly, at the cost of running a local embedding server.
