---
title: "Comparing strings with ditto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing strings with ditto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
#| message: false
library(ditto)
```

`ditto` compares two pieces of text and reduces the comparison to a single
number, so that comparisons can be ranked, filtered, or summarised at scale.
Typical uses are checking whether a survey question has been reworded across
waves, comparing a translation against its source, and grouping near-duplicate
free-text responses.

The package collects metrics that differ in what they treat as a meaningful
difference. The edit-distance, Jaccard, and cosine metrics come from
[`stringdist`](https://github.com/markvanderloo/stringdist) and compare
characters; `bleu()` compares word sequences; and the embedding metrics,
`bertscore()` and `cosine_similarity()`, compare meaning through contextual
embeddings. `compare_strings()` runs the character- and word-level metrics
together, and the embedding metrics too when an embedding server is available,
returning them in one table.

## Cleaning text

Two strings can differ only in capitalisation or punctuation while expressing
exactly the same thing. `clean()` removes those surface differences by
lowercasing, stripping punctuation, and collapsing repeated whitespace, so
that they do not lower the similarity scores.

```{r}
#| label: clean
clean("To what extent do you AGREE with the statement?")
```

The rest of this vignette compares a reference survey question against four
variants that range from a near-identical rewording to an unrelated question.
Each string is cleaned first.

```{r}
#| label: data
reference <- clean("To what extent do you agree with the following statements?")

variants <- c(
  "To what extent do you agree with the following statement?",
  "How much do you agree with each of the statements below?",
  "How satisfied are you with the service you received?",
  "What is your highest level of education completed?"
)
labels <- c("Near-identical", "Paraphrase", "Different", "Unrelated")
variants <- clean(variants)
```

## Character and word metrics

`compare_strings()` computes four metrics for each candidate against the
reference and returns a tibble with one row per pair. Three are character
metrics from `stringdist`: Levenshtein (single-character edits), Jaccard
(overlap of the character sets), and cosine (overlap of character
frequencies). The fourth is BLEU, which is described below.

```{r}
#| label: compare
result <- compare_strings(variants, reference)
cbind(label = labels, round(result[, -(1:2)], 2))
```

The near-identical variant scores near 1 on every metric. The other three
diverge in ways that reveal what each metric responds to. The unrelated
question still scores 0.92 on cosine -- here a *character* cosine, the overlap
of character frequencies, not to be confused with the meaning-based
`cosine_similarity()` introduced later -- because it happens to reuse many of
the same characters. None of these metrics compares meaning, so surface overlap
can make an unrelated string look similar.

## BLEU

`bleu()` evaluates a candidate against a reference at the level of word
sequences rather than characters. It counts how many of the candidate's
n-grams (single words, pairs, triples, and 4-word sequences) also appear in
the reference, credits each reference n-gram only as many times as it occurs,
and applies a penalty when the candidate is shorter than the reference. The
maximum n-gram order is capped at the length of the shorter string, so short
inputs are scored on the orders that are actually possible.

```{r}
#| label: bleu
bleu(
  "how much do you agree with the statement",
  "to what extent do you agree with the statement"
)
```

BLEU returns 0 when any n-gram order has no matches at all, which is why the
"Different" and "Unrelated" variants score 0 in the table above: neither
shares a 4-word sequence with the reference. This makes BLEU stricter than the
character metrics on word order, and well suited to translation evaluation,
where the arrangement of words carries meaning.

## Semantic similarity with embeddings

The metrics so far compare the characters or words that are present. A
paraphrase that conveys the same meaning with different words scores low on all
of them. The package's two embedding metrics compare *meaning* instead.
`bertscore()` embeds each token in the context of its sentence and matches every
candidate token to its most similar reference token, reporting precision,
recall, and their harmonic mean (F1). `cosine_similarity()` instead pools each
string into a single vector and takes the cosine between the two.

These are the only parts of the package that need an external model, because
they require one embedding *per token* rather than one pooled vector per string.

### Setting up the embedding server

`bertscore()` and `cosine_similarity()` read per-token embeddings from a local
[`llama.cpp`](https://github.com/ggml-org/llama.cpp) server. The friendly
local runners, Ollama and LM Studio, return only a single pooled vector per
string through their OpenAI-style endpoints, so they cannot supply the
token-level embeddings these metrics need. A `llama.cpp` server started with
`--pooling none` can.

1. Download the prebuilt `llama.cpp` binaries for your platform from the
   [releases page](https://github.com/ggml-org/llama.cpp/releases) (no
   compilation required).
2. Download an embedding model in GGUF format. This vignette uses
   [`bge-m3`](https://huggingface.co/BAAI/bge-m3), a multilingual model that
   needs no task prefix; `all-MiniLM-L6-v2` is a smaller, English-only
   alternative. Some models, such as the E5 family, require a task prefix like
   `"query: "`; pass it through the `prefix` argument so its tokens are stripped
   before matching, rather than treated as part of the text.
3. Start the server with embeddings and pooling disabled:

```sh
llama-server -m bge-m3-f16.gguf --embedding --pooling none --port 8080
```

The default `host` argument points at `http://localhost:8080`; pass a
different URL if the server runs elsewhere.

Rather than start the server by hand, `start_llama_server()` launches it for
you with the right flags and waits until it is ready to answer requests.
`stop_llama_server()` shuts it down again, and `llama_server_running()` reports
whether a server is responding. Give it the paths to your `llama-server`
binary and `.gguf` model; `exe` can be omitted if `llama-server` is on your
`PATH`.

```{r}
#| label: start-server
#| eval: false
start_llama_server(
  model = "path/to/bge-m3-f16.gguf",
  exe = "path/to/llama-server"
)
llama_server_running()
```

If you set the binary and model paths once as options in your `.Rprofile`,
`start_llama_server()` needs no arguments:

```{r}
#| label: server-options
#| eval: false
options(
  ditto.llama_server = "path/to/llama-server",
  ditto.llama_model = "path/to/bge-m3-f16.gguf"
)
start_llama_server()
```

### Computing a BERTScore

With the server running, `bertscore()` returns a named vector of precision,
recall, and F1. The examples that follow are not run when this vignette is
built, since they depend on the server; the output shown is from a live run
against `bge-m3`.

```{r}
#| label: bertscore
#| eval: false
bertscore("how much do you agree", "to what extent do you agree")
#>  precision     recall         f1
#>      0.944      0.938      0.941

bertscore("what is your date of birth", "to what extent do you agree")
#>  precision     recall         f1
#>      0.709      0.720      0.714
```

The paraphrase scores 0.94 and the unrelated question 0.71. Both are high:
`bge-m3` places even unrelated text at a substantial similarity, so the raw
separation is narrow. The ordering is correct, but these scores need calibrating
before they are comfortable as thresholds; *Calibrating scores with a baseline*
below does exactly that. `token_embeddings()` exposes the underlying matrix if
you want to inspect it directly; it has one row per token, after dropping the
model's special tokens, and one column per embedding dimension.

```{r}
#| label: token-embeddings
#| eval: false
dim(token_embeddings("to what extent do you agree"))
#> [1]    6 1024
```

### Whole-string cosine

`cosine_similarity()` is the pooled counterpart to `bertscore()`: instead of
matching tokens, it collapses each string into a single vector and takes the
cosine between them. It is faster and is the classic sentence-embedding
similarity. Because the server returns per-token vectors, the pooling into one
vector happens in R, controlled by `pooling`. `"mean"` averages the token
vectors, as sentence-transformers models like `all-MiniLM` and the E5 family do;
`"cls"` takes the leading special token's vector, as BGE models like `bge-m3` do.
Use the pooling your model was trained with.

```{r}
#| label: cosine
#| eval: false
# bge-m3 is a CLS-pooling model.
cosine_similarity("how much do you agree", "to what extent do you agree",
                  pooling = "cls")
#> [1] 0.892
cosine_similarity("what is your date of birth", "to what extent do you agree",
                  pooling = "cls")
#> [1] 0.474
```

With the matching CLS pooling the unrelated pair drops to 0.47 while the
paraphrase stays at 0.89, a wider gap than mean pooling gives (0.72 versus 0.95),
because the mean blurs a sentence into an average that even unrelated questions
share. This `cosine_similarity()` compares meaning and should not be confused
with the character-frequency `cosine` column of `compare_strings()`.

### Matching across languages

Because `bge-m3` is multilingual, the embedding metrics recognise a paraphrase
even across languages, where the character and word metrics see no overlap at
all. Comparing the English reference with its Dutch translation:

```{r}
#| label: multilingual
#| eval: false
bertscore("to what extent do you agree", "in hoeverre bent u het eens")[["f1"]]
#> [1] 0.921
cosine_similarity("to what extent do you agree", "in hoeverre bent u het eens",
                  pooling = "cls")
#> [1] 0.895
```

Both score the cross-lingual pair almost as high as an English paraphrase.

### All metrics together

Setting `bert = TRUE` adds the two embedding columns, `bertscore_f1` and
`cosine_emb`, to `compare_strings()`; pass `pooling = "cls"` for `bge-m3`. The
full table makes the difference between surface and semantic similarity visible.

```{r}
#| label: compare-bert
#| eval: false
compare_strings(variants, reference, bert = TRUE, pooling = "cls")
```

| label          | levenshtein | jaccard | cosine | bleu | bertscore_f1 | cosine_emb |
|:---------------|------------:|--------:|-------:|-----:|-------------:|-----------:|
| Near-identical |        0.98 |    1.00 |   1.00 | 0.88 |         0.99 |       0.97 |
| Paraphrase     |        0.49 |    0.86 |   0.94 | 0.26 |         0.96 |       0.93 |
| Different      |        0.35 |    0.67 |   0.84 | 0.00 |         0.78 |       0.65 |
| Unrelated      |        0.25 |    0.82 |   0.92 | 0.00 |         0.71 |       0.49 |

Both embedding metrics keep the paraphrase high (0.96 and 0.93) even though its
BLEU is low (0.26), because it preserves meaning while changing words. The
different and unrelated questions, which character cosine left at a misleading
0.84 and 0.92, separate much better on `cosine_emb` (0.65 and 0.49) than on the
anisotropic `bertscore_f1` (0.78 and 0.71) -- though the baseline rescaling below
recovers that separation for BERTScore too.

### Calibrating scores with a baseline

Raw BERTScore values are not comparable across models, and some models compress
the range. With `all-MiniLM-L6-v2` an unrelated pair scores around 0.11, but
with a model whose token embeddings are anisotropic, such as the multilingual
`bge-m3`, *unrelated* text already scores around 0.71. The ordering is still
correct -- paraphrases score higher than unrelated pairs -- but the usable range
is squeezed into a narrow band near the top, and a fixed threshold means
something different for each model.

The original BERTScore addresses this by rescaling against a baseline: the score
the model assigns to unrelated text. `bertscore_baseline()` estimates it by
averaging over many random, and therefore unrelated, pairs drawn from a corpus.
By default it uses the English rows of `example_sentences`, a bundled set of
distinct-meaning sentences; pass your own `texts` for a domain- or
language-specific estimate.

```{r}
#| label: baseline
#| eval: false
baseline <- bertscore_baseline(seed = 1)
baseline
#>  precision     recall         f1
#>      0.628      0.629      0.628
```

Passing this `baseline` to `bertscore()` (or `compare_strings()`) rescales each
score as `(x - baseline) / (1 - baseline)`, mapping the unrelated-text floor to
0 while leaving a perfect match at 1.

```{r}
#| label: baseline-rescale
#| eval: false
# Raw scores from bge-m3: paraphrase and unrelated sit close together.
bertscore("how much do you agree", "to what extent do you agree")[["f1"]]
#> [1] 0.941
bertscore("what is your date of birth", "to what extent do you agree")[["f1"]]
#> [1] 0.714

# Rescaled against the baseline: the gap opens up.
bertscore(
  "how much do you agree",
  "to what extent do you agree",
  baseline = baseline
)[["f1"]]
#> [1] 0.842
bertscore(
  "what is your date of birth",
  "to what extent do you agree",
  baseline = baseline
)[["f1"]]
#> [1] 0.231
```

The baseline is specific to the model and the language, so estimate it from text
that matches your use. For a multilingual model, `bertscore_baselines()` returns
one baseline per language from a labelled corpus (`example_sentences` by
default), so each language can be rescaled against its own floor:

```{r}
#| label: baseline-language
#| eval: false
baselines <- bertscore_baselines(seed = 1)
bertscore("in hoeverre bent u het eens", "hoeveel bent u het eens",
          baseline = baselines[["nl"]])
```

## Choosing a metric

The metrics answer different questions, so the choice follows from what counts
as a difference.

- Use the **character metrics** when responses should closely match the
  reference and you expect only typos, small edits, or reordering. They are
  fast and need no external model. Cosine is usually the most informative of
  the three; Jaccard adds little beside it.
- Use **BLEU** when word order matters, such as translation evaluation, and
  you want a candidate penalised for rearranging or dropping words.
- Use the **embedding metrics** when paraphrasing is expected and valid
  responses may use different words, or even different languages, for the same
  meaning. They are the only metrics here that score such responses correctly,
  at the cost of running a local embedding server. `cosine_similarity()` is
  faster and gives one number per pair; `bertscore()` is finer-grained, matching
  token by token, and reports precision and recall separately.

Two practical notes for the embedding metrics: match the `pooling` argument to
your model (`"cls"` for `bge-m3`, `"mean"` for `all-MiniLM` and the E5 family),
and rescale `bertscore()` against a baseline when you need thresholds that are
comparable across models.
