---
title: "The metrics, and what they are validated against"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The metrics, and what they are validated against}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
#| message: false
library(ditto)
```

A similarity metric is only useful if it computes what its name says it does.
BLEU, CHRF, ROUGE, TER, and METEOR are all published metrics with widely used
reference implementations, so `ditto`'s versions can be checked against them
rather than merely tested for plausibility. This vignette records what each
metric is checked against, under what settings, and where `ditto` knowingly
departs from the reference.

The short answer: every metric `ditto` implements from scratch is validated
against an external implementation, and the departures are deliberate, few, and
listed below.

## What ditto computes

Three of the columns `compare_strings()` returns are not `ditto`'s own work at
all: `levenshtein`, `jaccard`, and `cosine` come from
[`stringdist`](https://github.com/markvanderloo/stringdist), which is where the
question of validating them belongs. The rest are implemented here.

```{r}
#| label: metric-table
#| echo: false
metrics <- tibble::tribble(
  ~Metric,            ~Function,               ~`Validated against`,
  "BLEU",             "bleu()",                "sacrebleu",
  "CHRF",             "chrf()",                "sacrebleu",
  "ROUGE-1/2/L",      "rouge()",               "rouge-score (Google)",
  "TER",              "ter()",                 "sacrebleu",
  "Word error rate",  "wer()",                 "jiwer",
  "METEOR",           "meteor()",              "nltk",
  "BERTScore",        "bertscore()",           "bert-score (original authors)",
  "Embedding cosine", "cosine_similarity()",   "not validated",
  "Levenshtein, Jaccard, character cosine", "compare_strings()", "delegated to stringdist"
)
knitr::kable(metrics)
```

`cosine_similarity()` is the one gap. It is the cosine between two pooled
embedding vectors, which has no reference implementation to disagree with: once
the embeddings and the pooling are fixed, the arithmetic is a one-liner. The
embeddings themselves are exercised through `bertscore()`, which *is* validated.

## What has to match

Two implementations agree only when they are asked for the same quantity. Each
reference below is configured to compute exactly what `ditto` computes; these
are the settings that matter.

**Tokenization**, first, because it underlies everything else. `ditto`'s word
metrics split on whitespace and peel punctuation into its own token, following
TER's original convention and `sacrebleu`'s `13a` tokenizer. Every reference is
either handed text already tokenized this way, or told to use that tokenizer, so
no disagreement can be blamed on tokenization.

`bleu()` is checked against `sacrebleu`'s `BLEU` with `smooth_method="none"`
(`ditto` applies no smoothing), `effective_order=True`, and `tokenize="none"`
so `sacrebleu` accepts the pre-tokenized text.

`chrf()` is checked against `sacrebleu`'s `CHRF` with `char_order=6`, `beta=2`,
`word_order=0` (plain CHRF, not chrF++), `whitespace=False`, and no epsilon
smoothing. These are `sacrebleu`'s defaults and `ditto`'s.

`rouge()` is checked against Google's `rouge_score` with `use_stemmer=False`,
since `rouge()` does not stem, and its F-measure corresponds to `ditto`'s
default `beta = 1`.

`ter()` is checked against `sacrebleu`'s `TER` with `case_sensitive=True`,
because `ditto` compares the text as given while `sacrebleu` lowercases by
default. With `normalized=False` and `no_punct=False`, `sacrebleu`'s tokenizer
then only collapses whitespace, so the pre-tokenized text passes through
untouched.

`wer()` is checked against `jiwer`'s word error rate on the same tokens. It is
`ter()` without the shift search, so word error rate is the right reference for
it, and the two functions bracket what a reordering ought to cost.

`meteor()` is checked against `nltk` with the Snowball English stemmer (the
stemmer behind `SnowballC::wordStem()`, not `nltk`'s default Porter stemmer), a
stubbed WordNet that disables the synonym stage `ditto` omits, and
`alpha=0.9, beta=3, gamma=0.5`, which is exactly `ditto`'s
`f_mean = 10PR / (R + 9P)` and `penalty = 0.5 * (chunks / matches)^3`.

`bertscore()` has its own vignette, `vignette("bertscore-validation")`, because
it is the only metric that needs a model server on both sides.

## The comparison

The scores below were produced by the reference implementations under those
settings. `ditto` recomputes the same pairs when this vignette is built, so the
agreement is checked afresh rather than asserted.

```{r}
#| label: reference-scores
reference <- tibble::tribble(
  ~candidate, ~reference, ~bleu, ~chrf, ~rouge_1, ~rouge_2, ~rouge_l, ~ter, ~wer, ~meteor,
  "the cat sat on the mat", "the cat sat on the mat",
  1.0000000, 1.0000000, 1.0000000, 1.0000000, 1.0000000, 0.0000000, 0.0000000, 0.9976852,
  "the cat sat on the mat", "a cat was sitting on the mat",
  0.0000000, 0.3722543, 0.6153846, 0.3636364, 0.6153846, 0.4285714, 0.4285714, 0.5434783,
  "how much do you agree with the statement", "to what extent do you agree with the statement",
  0.6004288, 0.6985244, 0.7058824, 0.6666667, 0.7058824, 0.3333333, 0.3333333, 0.6725968,
  "b a c d", "a b c d",
  0.0000000, 0.3333333, 1.0000000, 0.3333333, 0.7500000, 0.2500000, 0.5000000, 0.7890625,
  "d e f a b c", "a b c d e f",
  0.0000000, 0.3833333, 1.0000000, 0.8000000, 0.5000000, 0.1666667, 1.0000000, 0.9814815,
  "she walked slowly to the store", "slowly she walked to the shop",
  0.0000000, 0.6141196, 0.8333333, 0.4000000, 0.6666667, 0.3333333, 0.5000000, 0.7433333,
  "i went the market to buy bread", "i went to the market to buy bread",
  0.6129752, 0.8530376, 0.9333333, 0.7692308, 0.9333333, 0.1250000, 0.1250000, 0.8757427,
  "running runs runner", "run running ran",
  0.0000000, 0.5631993, 0.3333333, 0.0000000, 0.3333333, 1.0000000, 1.0000000, 0.3333333,
  "one two three four five", "five four three two one",
  0.0000000, 0.5073802, 1.0000000, 0.0000000, 0.2000000, 0.8000000, 0.8000000, 0.5000000,
  "do you agree with this statement.", "do you agree with this statement?",
  0.8091067, 0.9606070, 0.8571429, 0.8333333, 0.8571429, 0.1428571, 0.1428571, 0.8551587
)
```

```{r}
#| label: ditto-scores
pairwise <- function(f, ...) {
  mapply(f, reference$candidate, reference$reference,
    MoreArgs = list(...), USE.NAMES = FALSE
  )
}

ours <- tibble::tibble(
  bleu = pairwise(bleu),
  chrf = pairwise(chrf),
  rouge_1 = pairwise(rouge, variant = "1"),
  rouge_2 = pairwise(rouge, variant = "2"),
  rouge_l = pairwise(rouge, variant = "l"),
  ter = pairwise(ter),
  wer = pairwise(wer),
  meteor = pairwise(meteor)
)
```

The largest disagreement on any pair, for each metric:

```{r}
#| label: agreement
#| echo: false
theirs <- reference[names(ours)]
largest <- vapply(names(ours), function(m) max(abs(ours[[m]] - theirs[[m]])), numeric(1))

knitr::kable(
  tibble::tibble(
    metric = names(largest),
    `largest difference` = format(largest, digits = 2, scientific = TRUE)
  )
)
```

Every metric agrees to within about `1e-7`, which is the rounding in the table
above rather than a real difference; the validation scripts, which carry full
precision, agree to about `1e-11`. The scores themselves are worth a glance too,
because they show the metrics disagreeing with each other in exactly the ways
they are designed to:

```{r}
#| label: score-table
#| echo: false
knitr::kable(cbind(pair = seq_len(nrow(ours)), round(ours, 3)))
```

Pair 2 is a paraphrase that shares no 4-gram, so BLEU is 0 while ROUGE-1 is
0.62. Pair 5 (`d e f a b c` against `a b c d e f`) is one block of three words
moved: TER charges a single shift for 0.17, while word error rate rewrites every
word for 1.0. Pair 8 (`running runs runner`) matches only by stem, so BLEU and
ROUGE-2 are 0 while METEOR is 0.33. Pair 9 is a pure reversal: every word is
present, so ROUGE-1 is a perfect 1.0, while ROUGE-L collapses to 0.2 and both
error rates climb to 0.8 — no contiguous block can be moved to fix a reversal.
A single metric would have hidden each of these.

## TER and word error rate

TER counts a contiguous block of words moved to a new position as a single
edit, a *shift*. Finding the cheapest set of shifts is NP-complete (Shapira &
Storer, 2002), so TER's original implementation, TERCOM, searches greedily:
apply the shift that most reduces the edit distance, repeat until none helps.
`ter()` reproduces that search, closely enough to return `sacrebleu`'s score.

`wer()` is the same edit rate without shifts. Together they bracket what a
reordering costs, which is often the more informative reading:

```{r}
#| label: ter-vs-wer
c(ter = ter("she walked slowly to the store", "slowly she walked to the shop"),
  wer = wer("she walked slowly to the store", "slowly she walked to the shop"))
```

Moving `"slowly"` is one edit for TER and two for word error rate. `ter()` costs
more to compute, since the shift search rescores the sentence for every
candidate shift; on sentence-length inputs this is milliseconds, but it is not
free the way `wer()` is.

## Known departures

Two, both deliberate.

**`bleu()` caps its n-gram order at the shorter string.** `sacrebleu`'s
`effective_order` caps at the *candidate's* length. The two coincide unless the
reference is the shorter side and has fewer than four tokens, where `sacrebleu`
scores the impossible orders as 0 and returns 0, while `ditto` evaluates the
orders that are actually possible and returns a non-zero score. This keeps short
survey items from all collapsing to 0, at the cost of departing from
`sacrebleu` there.

```{r}
#| label: bleu-departure
bleu("a b c d e", "a b") # sacrebleu returns 0
bleu("a b", "a b c d e") # candidate is shorter; both return this
```

**`meteor()` has no synonym stage.** METEOR's third matching stage uses WordNet,
which exists for only a handful of languages. `ditto` covers exact and stem
matching, so it agrees with `nltk` on text where no synonyms would have matched
and scores lower where they would have. Its alignment is also a heuristic, not
the chunk-minimal alignment the original tie-break calls for; it reproduces
`nltk`'s right-to-left scan so the two agree, but a left-to-right scan would
sometimes find fewer chunks.

`chrf()`, `rouge()`, `ter()`, and `wer()` have no known departure. `chrf()`
agrees with `sacrebleu` even on strings shorter than the n-gram order, where
`ditto`'s cap coincides with `sacrebleu`'s effective-order handling. `ter()`
agrees with `sacrebleu` on sentences long enough that its beam search prunes
the edit-distance matrix, not only on short ones where the beam is inactive.

## Reproducing this

The comparison is not run when the vignette is built, since it needs Python.
Two scripts in `dev/validation/` regenerate it:

```sh
pip install sacrebleu rouge-score jiwer nltk
python dev/validation/text-metrics-reference.py   # writes the reference scores
Rscript dev/validation/text-metrics-ditto.R       # scores the same pairs, diffs them
```

The R script stops with an error if any validated metric drifts from its
reference, and reports how often `ter()` beats `wer()`, so a silently broken
shift search cannot pass as word error rate. `bertscore-reference.py` and
`bertscore-ditto.R` do the same for BERTScore, which additionally needs a
running `llama.cpp` server.

## Other implementations

Worth knowing what else exists, in R and elsewhere.

**In R the coverage is thin.**
[`sacRebleu`](https://cran.r-project.org/package=sacRebleu) is the only CRAN
package covering any of these metrics, and despite its name it implements BLEU
only, in C++, with the smoothing methods of Chen and Cherry (2014).
[`rougeR`](https://github.com/LJCovingtonJr/rougeR) is GitHub-only and covers
ROUGE-N and ROUGE-S, with ROUGE-L listed as in development.
[`wersim`](https://rdrr.io/github/jenswaeckerle/wersim/man/wer.html) is
GitHub-only and computes word error rate. The
[`text`](https://cran.r-project.org/package=text) package gives you transformer
embeddings and a `textSimilarity()`, but not BERTScore's token-matching
formulation. No CRAN package appears to implement CHRF, METEOR, TER, or
BERTScore, which is the gap `ditto` fills.

**In Python they are all well covered**, which is what made the validation
possible.

```{r}
#| label: python-table
#| echo: false
knitr::kable(tibble::tribble(
  ~Library,            ~Covers,
  "sacrebleu",         "BLEU, CHRF, chrF++, TER",
  "nltk",              "BLEU, METEOR, RIBES",
  "rouge-score",       "ROUGE-1/2/L/Lsum",
  "jiwer",             "WER, CER",
  "torchmetrics.text", "BLEU, ROUGE, CHRF, TER, WER, BERTScore",
  "evaluate",          "wrappers around most of the above",
  "bert-score",        "BERTScore"
))
```

One caveat on METEOR. `nltk`, which `ditto` matches score for score, is itself a
simplification of the official Java implementation of Denkowski and Lavie, which
performs a real alignment search and tunes its parameters per language. Matching
`nltk` buys agreement with what people actually run, not with the paper.

## References

Papineni, K., Roukos, S., Ward, T., & Zhu, W.-J. (2002). BLEU: a method for
automatic evaluation of machine translation. *Proceedings of the 40th Annual
Meeting of the Association for Computational Linguistics*, 311-318.

Popović, M. (2015). chrF: character n-gram F-score for automatic MT evaluation.
*Proceedings of the Tenth Workshop on Statistical Machine Translation*, 392-395.

Lin, C.-Y. (2004). ROUGE: A package for automatic evaluation of summaries.
*Text Summarization Branches Out*, 74-81.

Snover, M., Dorr, B., Schwartz, R., Micciulla, L., & Makhoul, J. (2006). A study
of translation edit rate with targeted human annotation. *Proceedings of the 7th
Conference of the Association for Machine Translation in the Americas*, 223-231.

Shapira, D., & Storer, J. A. (2002). Edit distance with move operations.
*Proceedings of the 13th Annual Symposium on Combinatorial Pattern Matching*,
85-98.

Banerjee, S., & Lavie, A. (2005). METEOR: An automatic metric for MT evaluation
with improved correlation with human judgments. *Proceedings of the ACL Workshop
on Intrinsic and Extrinsic Evaluation Measures for Machine Translation and/or
Summarization*, 65-72.

Zhang, T., Kishore, V., Wu, F., Weinberger, K. Q., & Artzi, Y. (2020).
BERTScore: Evaluating text generation with BERT. *International Conference on
Learning Representations (ICLR)*.
